---
title: "Solution - Lab 5: Getting data into R"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<script>
  var password = "MAT186LABSOER";
  var res = prompt("DCC Instructor Key Required:");
  if (res != password) {
    alert("Incorrect password. Access denied.");
    window.location.href = "../index.html";
  }
</script>

## Getting data into R - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Downloading the Dataset
Students were required to download the `lbw.csv` dataset. Success in this task is confirmed by the file appearing in the project directory.

### Task 2: Importing the Data
To import the data correctly using the Tidyverse, the `library(tidyverse)` must be called first. We use `read_csv()` to create a tibble.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Task 2: Import the data
# Assumes 'lbw.csv' is in the same folder as this .qmd file
lbw_data <- read_csv("../data/lbw.csv")
```

# Task 3: Show the structure
```{r}
glimpse(lbw_data)
```

# Task 4: Display the first 6 rows of the table
```{r}
head(lbw_data)
```



<br><br>

***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::

