---
title: "Solution - Lab 7: Cleaning Data"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<script>
  var password = "MAT186LABSOER";
  var res = prompt("DCC Instructor Key Required:");
  if (res != password) {
    alert("Incorrect password. Access denied.");
    window.location.href = "../index.html";
  }
</script>

## Cleaning Data - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Introduction to Data Cleaning
In this lab, we move from simply looking at data to **transforming** it. Raw data from portals like NY Open Data often contains too many variables or messy formatting. We use the `dplyr` library to keep only what is relevant to our research question.



---

### Task 2: Loading the Dataset
We continue using the `tidyverse` package. We must be careful with our file paths; since this solution is in the `solutions/` folder, we use `../` to find our data.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Load the SUNY Enrollment data
ny_data <- read_csv("../data/race_ny.csv")
```


### Task 3-5: The Cleaning Pipeline
The goal of this task was to create a "Clean" version of our table by selecting 4 key columns, filtering for a specific sector, and renaming columns for better readability.

#### Our Cleaning Strategy:

* **Select:** We kept Term , Institution Sector, Total Enrollment, and Hispanic/Latino.

* **Rename:** We shortened the names to semester, sector, total, and hispanic_latino.

* **Filter:** We narrowed the results to only show **"Community Colleges"**.

```{r}
#| label: data-cleaning-steps

# Create the cleaned object
cleaned_suny <- ny_data |> 
  # Task 2: Select the 4 most important columns by their position
  # This avoids "Column not found" errors caused by hidden spaces!
  select(1, 2, 3, 6) |> 
  
  # Task 4: Rename the columns for clarity
  # rename(new_name = old_position_or_name)
  rename(semester = 1,
         sector = 2,
         total_enrollment = 3,
         hispanic_latino = 4) |> 
  
  # Task 3: Filter for a specific category
  filter(sector == "Community Colleges")

# Task 5: Show the first 10 rows
head(cleaned_suny, n = 10)
```


#### Narrative Description of Results

By applying these transformations, we have turned a large dataset into a focused report.

* The Observation: Each row now represents a specific semester's enrollment data for the Community College sector.

* The Benefit: Instead of 12 columns, we now have the 4 most important variables for analyzing Hispanic/Latino enrollment trends in community colleges.


<br><br>

***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::

