---
title: "Solution - Lab 5: Getting data into R"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<div id="password-gate" style="background: #f8f9fa; padding: 20px; border: 1px solid #dee2e6; border-radius: 5px; text-align: center;">
  <h3>🔒 Solutions Protected</h3>
  <p>Please enter the classroom code to unlock Lab results.</p>
  <input type="password" id="pass-input" style="padding: 5px;" placeholder="Enter password...">
  <button onclick="checkPassword()" style="padding: 5px 15px; background: #007bff; color: white; border: none; border-radius: 3px; cursor: pointer;">Unlock</button>
</div>

<div id="solution-content" style="display:none;">

## Getting data into R - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Reading the Dataset from a URL

Students were required to download the `lbw.csv` dataset. Success in this task is confirmed by the file appearing in the project directory.


### Task 2: Downloading the Data from a URL


### Task 3: Importing the Dataset into RStudio
To import the data correctly using the Tidyverse, the `library(tidyverse)` must be called first. We use `read_csv()` to create a tibble.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Task 2: Import the data
# Assumes 'lbw.csv' is in the same folder as this .qmd file
lbw_data <- read_csv("../data/lbw.csv")
```

### Task 4: Show the structure
```{r}
glimpse(lbw_data)
```

### Task 5: Display the first 6 rows of the table
```{r}
head(lbw_data)
```



<br><br>

</div>

<script>
function checkPassword() {
var userPass = document.getElementById("pass-input").value;

// This uses the LAB_PASSWORD from your security.js file
if (userPass === LAB_PASSWORD) {
document.getElementById("password-gate").style.display = "none";
document.getElementById("solution-content").style.display = "block";
} else {
alert("Incorrect password. Please try again.");
}
}

// Allow "Enter" key to trigger the button
document.getElementById("pass-input").addEventListener("keypress", function(event) {
if (event.key === "Enter") {
checkPassword();
}
});
</script>



***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::

