---
title: "Solution - Lab 8: Exploratory Data Analysis"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<script>
  var password = "MAT186LABSOER";
  var res = prompt("DCC Instructor Key Required:");
  if (res != password) {
    alert("Incorrect password. Access denied.");
    window.location.href = "../index.html";
  }
</script>

## Exploratory Data Analysis - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Preparing the Data
We load our library and re-apply the cleaning steps from Lab 7 to ensure we are working with the correct variables.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# 1. Load data
ny_data <- read_csv("../data/race_ny.csv")

# 2. Re-apply Lab 7 Cleaning (using indices for accessibility)
cleaned_suny <- ny_data |> 
  select(1, 2, 3, 6) |> 
  rename(semester = 1,
         sector = 2,
         total_enrollment = 3,
         hispanic_latino = 4) |> 
  filter(sector == "Community Colleges")
```

### Task 2: Mean and Standard Deviation
We calculate the summary statistics for our main numeric variable: total_enrollment.

```{r}
#| label: summary-stats
stats <- cleaned_suny |> 
  summarise(
    mean_enrollment = mean(total_enrollment, na.rm = TRUE),
    sd_enrollment = sd(total_enrollment, na.rm = TRUE)
  )

stats
```

### Task 3: Histogram
A histogram helps us see the distribution of enrollment across different community college campuses and semesters.

```{r}
#| label: histogram
ggplot(cleaned_suny, aes(x = total_enrollment)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 15) +
  labs(title = "Distribution of Total Enrollment",
       subtitle = "SUNY Community Colleges",
       x = "Total Students",
       y = "Count") +
  theme_minimal()
```


**Observation:** The data appears to be right-skewed. Most community colleges have a smaller enrollment (under 10,000), while a few very large institutions pull the tail to the right.

### Task 4: Scatter Plot with Trend Line
We compare total_enrollment to hispanic_latino enrollment to see if they grow together.

```{r}
#| label: scatter-plot
ggplot(cleaned_suny, aes(x = total_enrollment, y = hispanic_latino)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Total Enrollment vs. Hispanic/Latino Enrollment",
       x = "Total Students",
       y = "Hispanic/Latino Students") +
  theme_minimal()
```
### Task 5: Boxplot by Category
We compare the distribution of total enrollment across different semesters.
```{r}
#| label: boxplot
ggplot(cleaned_suny, aes(x = semester, y = total_enrollment)) +
  geom_boxplot(fill = "orange", alpha = 0.7) +
  labs(title = "Enrollment Distribution by Semester",
       x = "Semester",
       y = "Total Students") +
  theme_minimal() +
  coord_flip() # Flipping for easier reading of labels
```

**Final Observation (3 Sentences)**
In these plots, there is a very strong positive correlation between the size of a college and its Hispanic/Latino student population, as shown by the linear trend in the scatter plot. Interestingly, the boxplot reveals that total enrollment has been gradually declining across the semesters, evidenced by the median line shifting downward. A major outlier exists in the community college sector, representing a massive campus that significantly exceeds the average enrollment of all other schools.




<br><br>

***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::

