---
title: "Solution - Lab 9: Probability distribution"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<script>
  var password = "MAT186LABSOER";
  var res = prompt("DCC Instructor Key Required:");
  if (res != password) {
    alert("Incorrect password. Access denied.");
    window.location.href = "../index.html";
  }
</script>

## Probability distribution - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Accessible Data Loading
We load the `tidyverse` and import our dataset. Because our CSV uses commas inside numbers (e.g., "192,959"), we must clean the data so R can perform calculations.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Import the data
ny_data <- read_csv("../data/race_ny.csv")

# Data Cleaning for Accessibility:
# 1. Select 'Total Enrollment' (Column 3)
# 2. Remove commas so R sees it as a number
clean_data <- ny_data |>
  mutate(total_enroll = as.numeric(gsub(",", "", `Total Enrollment`))) |>
  filter(!is.na(total_enroll))
```

### Task 2: Distribution Visualization
We create a histogram of total_enrollment to check for normality.

```{r}
#| label: fig-histogram
#| fig-cap: "Histogram showing the frequency distribution of total student enrollment across SUNY sectors."
#| fig-alt: "A histogram with a long right tail showing that most SUNY sectors have lower enrollment, while a few have very high enrollment, indicating a non-normal, right-skewed distribution."
ggplot(clean_data, aes(x = total_enroll)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 20) +
  labs(title = "Distribution of SUNY Enrollment",
       x = "Total Students",
       y = "Frequency") +
  theme_minimal()
```

**Observation:** The data is not Normal. It is heavily skewed to the right. A "Normal" distribution would be symmetrical like a bell, but here most data is clustered on the left with a long tail on the right.

### Task 3: Center and Spread
We find the mean (the center) and the standard deviation (the spread) for our variable.

```{r}
#| label: summary-stats
e_mean <- mean(clean_data$total_enroll)
e_sd <- sd(clean_data$total_enroll)

# Display results in an accessible table
tibble(Metric = c("Mean (Center)", "Standard Deviation (Spread)"), 
       Value = c(round(e_mean, 2), round(e_sd, 2)))
```


### Task 4: The Central Limit Theorem Simulation
The Central Limit Theorem (CLT) tells us that if we take the average of many samples, that average will look normal, even if the original data does not.

**Small Sample Size (n = 5)**

```{r}
#| label: fig-sim5
#| fig-cap: "CLT Simulation with a small sample size (n = 5)."
#| fig-alt: "A histogram of 1,000 sample means where each sample size is 5. The distribution is somewhat spread out and slightly skewed."
set.seed(123)
sim_5 <- replicate(1000, mean(sample(clean_data$total_enroll, 5, replace = TRUE)))
hist(sim_5, col = "skyblue", border = "white", 
     main = "Sample Size n = 5", xlab = "Sample Means")
```

**Large Sample Size (n = 100)**

```{r}
#| label: fig-sim100
#| fig-cap: "CLT Simulation with a large sample size (n = 100)."
#| fig-alt: "A histogram of 1,000 sample means where each sample size is 100. The distribution is very narrow, tall, and perfectly bell-shaped."
set.seed(123)
sim_100 <- replicate(1000, mean(sample(clean_data$total_enroll, 100, replace = TRUE)))
hist(sim_100, col = "darkblue", border = "white", 
     main = "Sample Size n = 100", xlab = "Sample Means")
```

### Task 5: Observation

When the sample size increased from 5 to 100, the spread of the histogram became significantly narrower. The sample means became more concentrated around the true population mean, and the shape transformed into a clear, symmetrical bell curve. This demonstrates that larger samples provide more consistent and predictable estimates.

<br><br>

***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::

