---
title: "Solution - Lab 10: Statistical inference for one mean"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<div id="password-gate" style="background: #f8f9fa; padding: 20px; border: 1px solid #dee2e6; border-radius: 5px; text-align: center;">
  <h3>🔒 Solutions Protected</h3>
  <p>Please enter the classroom code to unlock Lab results.</p>
  <input type="password" id="pass-input" style="padding: 5px;" placeholder="Enter password...">
  <button onclick="checkPassword()" style="padding: 5px 15px; background: #007bff; color: white; border: none; border-radius: 3px; cursor: pointer;">Unlock</button>
</div>

<div id="solution-content" style="display:none;">

## Statistical inference for one mean - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Loading and Preparing Data
As in previous labs, we must clean the `Total Enrollment` column by removing commas and converting it to a numeric format to ensure the `t.test()` function can process the data correctly.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Load the data
ny_data <- read_csv("../data/race_ny.csv")

# Clean numeric variable for accessibility and analysis
clean_data <- ny_data |>
  mutate(total_enroll = as.numeric(gsub(",", "", `Total Enrollment`))) |>
  filter(!is.na(total_enroll))
```

### Task 2: State Your Hypotheses
We want to test if the true mean enrollment across all SUNY sectors is significantly different from 100,000 students.

* **Null Hypothesis ($H_0$):** The mean enrollment is equal to 100,000 ($\mu = 100,000$).
* **Alternative Hypothesis ($H_a$):** The mean enrollment is not equal to 100,000 ($\mu \neq 100,000$).

### Task 3: Running the T-Test
We use the t.test() function in R. We set mu = 100000 to test our hypothesis.

```{r}
#| label: t-test-results
#| fig-cap: "Summary of the one-sample t-test results for total enrollment."
# Perform the t-test
test_results <- t.test(clean_data$total_enroll, mu = 100000)

# Display the results
test_results
```

### Task 4: Visualizing the Distribution
To provide context for our test, we visualize the data distribution relative to our test value ($\mu$).
```{r}
#| label: fig-inference-plot
#| fig-cap: "Histogram of enrollment with the null hypothesis mean (100,000) marked as a dashed red line."
#| fig-alt: "A histogram showing SUNY enrollment distribution. A vertical dashed red line is positioned at 100,000, representing the null hypothesis value. The data appears spread out around this line."
ggplot(clean_data, aes(x = total_enroll)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 15) +
  geom_vline(xintercept = 100000, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Enrollment Distribution vs. Null Hypothesis",
       x = "Total Enrollment",
       y = "Frequency") +
  theme_minimal()
```
### Task 5: The Conclusion
Based on the results of our t-test:

**Formal Conclusion:** Since the p-value was r round(test_results$p.value, 4), which is greater than the standard alpha level of 0.05, we fail to reject the null hypothesis. There is not sufficient evidence to suggest that the mean enrollment is significantly different from 100,000.


<br><br>

</div>

<script>
function checkPassword() {
var userPass = document.getElementById("pass-input").value;

// This uses the LAB_PASSWORD from your security.js file
if (userPass === LAB_PASSWORD) {
document.getElementById("password-gate").style.display = "none";
document.getElementById("solution-content").style.display = "block";
} else {
alert("Incorrect password. Please try again.");
}
}

// Allow "Enter" key to trigger the button
document.getElementById("pass-input").addEventListener("keypress", function(event) {
if (event.key === "Enter") {
checkPassword();
}
});
</script>



***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::



