---
title: "Solution - Lab 11: Statistical inference for one proportion"
subtitle: "MAT 186: Introduction to Data Science"
format:
  html:
    toc: true
    toc-location: right
---

<script>
  var password = "MAT186LABSOER";
  var res = prompt("DCC Instructor Key Required:");
  if (res != password) {
    alert("Incorrect password. Access denied.");
    window.location.href = "../index.html";
  }
</script>

## Statistical inference for one proportion - Official Solutions

::: {.callout-note}
### Instructor Note
Add your specific DCC lab answers below this box.
:::
### Task 1: Loading and Counting Data
We load our dataset and count how many times "Community Colleges" appears compared to the total number of observations.

```{r}
#| message: false
#| warning: false
library(tidyverse)

# Load the data
ny_data <- read_csv("../data/race_ny.csv")

# 1. Count successes (Community Colleges)
successes <- ny_data |> 
  filter(`Institution Sector` == "Community Colleges") |> 
  nrow()

# 2. Count total observations
total_n <- nrow(ny_data)

# Display counts
tibble(Category = "Community Colleges", Successes = successes, Total = total_n)
```
### Task 2: State Your Hypotheses
We are testing the claim that exactly one-quarter of our data entries come from the Community College sector.Null Hypothesis ($H_0$): The proportion of Community Colleges is 0.25 ($p = 0.25$).Alternative Hypothesis ($H_a$): The proportion of Community Colleges is not 0.25 ($p \neq 0.25$).

### Task 3: Running the Proportion Test
We use prop.test() to determine if our sample proportion is significantly different from our hypothesized value of 0.25.

```{r}
#| label: prop-test-results
# Run the proportion test
# x = number of successes, n = total sample size, p = hypothesized proportion
prop_results <- prop.test(x = successes, n = total_n, p = 0.25)

# Display results
prop_results
```
### Task 4: Visualization for Accessibility
We visualize the frequency of each sector to show how the "Success" category compares to the others.
```{r}
#| label: fig-prop-viz
#| fig-cap: "Bar chart showing the frequency of different SUNY Institution Sectors in the dataset."
#| fig-alt: "A bar chart with four bars of equal height. Each sector (Community Colleges, Comprehensive Colleges, Doctoral, and Technology) has exactly 23 observations, representing 25% of the total data each."
ggplot(ny_data, aes(x = `Institution Sector`)) +
  geom_bar(fill = "darkred", color = "black") +
  labs(title = "Frequency of Institution Sectors",
       x = "Sector",
       y = "Number of Observations") +
  theme_minimal() +
  coord_flip()
```

### Task 5: Final Report
Based on the statistical output above, here is our formal summary:

* **Sample Proportion ($\hat{p}$):** The calculated sample proportion is r round(prop_results$estimate, 2).

* **Significance:** The p-value is r round(prop_results$p.value, 4). Since this is greater than 0.05, our result is not significant.

* **Confidence Interval:** We are 95% confident that the true proportion of Community Colleges in this dataset lies between r round(prop_results$conf.int[1], 3) and r round(prop_results$conf.int[2], 3).

* **Conclusion:** Since the p-value is 1 (or very high), we fail to reject the null hypothesis. There is no evidence to suggest the proportion is different from 0.25; in fact, it matches our hypothesis exactly.


<br><br>

***

::: {.content-block}
<center>
<a href="../solutions.html" class="btn btn-primary btn-lg" role="button">🏠 Back to Solutions Hub</a>
</center>
:::

