# Lab 7: Cleaning Data with dplyr

## Learning Objectives
* Understand the concept of "Data Wrangling."
* Use the **Pipe Operator** (`|>`) to chain commands.
* Learn core cleaning functions: `filter()`, `select()`, and `rename()`.
* Handle missing values (`NA`).

---

## Introduction
"Raw" data is rarely ready for analysis. It might have columns you don't need, rows that aren't relevant to your study, or names that are hard to type. In this lab, we use the `dplyr` package (part of the `tidyverse`) to "tidy up" our datasets.

---

## Part 1: The Pipe Operator (`|>`)
Think of the pipe operator as saying the words **"and then."** It takes the output of one function and passes it to the next.

Instead of writing:
`clean_data <- filter(raw_data, age > 18)`

We write:
`clean_data <- raw_data |> filter(age > 18)`



---

## Part 2: Core Cleaning Functions

### 1. `select()` - Picking Columns
If your dataset has 100 columns but you only need 3, use `select()`.

```{r}
#| label: tbl-starwars-heights
#| tbl-cap: "Table 1: Name and height of the first six characters in the Star Wars dataset."
#| message: false

library(tidyverse)

# Keep only the 'name' and 'height' columns
sw_small <- starwars |> 
  select(name, height)

# Use knitr::kable to generate an accessible HTML table
knitr::kable(head(sw_small))
```

### 2. `filter()` - Picking Rows
If you only want to look at specific observations (e.g., characters taller than 100cm).

```{r}
#| label: tbl-tall-starwars
#| tbl-cap: "Table: Subset of Star Wars characters with a height greater than 100cm."
#| message: false

library(tidyverse)

# Keep only rows where height is greater than 100
tall_characters <- starwars |> 
  filter(height > 100) |>
  select(name, height, species, homeworld) # Selecting specific columns makes tables easier to read

# Use kable to create an accessible HTML table structure
knitr::kable(head(tall_characters))
```

### 3. `rename()` - Changing Names
If a column name is confusing or has spaces, rename it.

```{r}
#| label: tbl-rename-homeworld
#| tbl-cap: "Table: Star Wars character data showing the renamed 'home_planet' column."
#| message: false

library(tidyverse)

# Change 'homeworld' to 'home_planet'
sw_clean <- starwars |> 
  rename(home_planet = homeworld)

# We select a few columns so the table isn't too wide for a screen reader
knitr::kable(head(sw_clean[, c("name", "height", "home_planet")]))
```

---

## Part 3: Handling Missing Data (`NA`)
In R, missing values are represented as `NA` (Not Available). You can't calculate the mean of a column if it has `NA` values without telling R what to do.

To remove rows with *any* missing data:
```{r}
#| label: tbl-drop-na
#| tbl-cap: "Table: Star Wars characters after removing rows with missing height values."
#| message: false

library(tidyverse)

# Remove rows where height is NA
tidy_sw <- starwars |> 
  drop_na(height)

# Display the first six rows using an accessible HTML table format
knitr::kable(head(tidy_sw[, c("name", "height", "mass", "species")]))
```



---

<div class="dcc-important">
**Lab Task 7:**

1. Open your Quarto document from Lab 6 (the one with your NY State Open Data).
2. Create a new section called "Data Cleaning."
3. Use `select()` to keep only the 4 most important columns in your dataset.
4. Use `filter()` to narrow your data down to one specific category or a certain range of numbers.
5. Use `rename()` to make at least one column name easier to read.
6. **Render** your document and check that your new "Cleaned" table looks correct.
</div>

---

## Pro-Tip: The "Glimpse"
Always run `glimpse(your_data)` after cleaning to make sure you haven't accidentally deleted something you meant to keep!

---

::: {.callout-tip}
### Done with the Lab?
If you have finished all the tasks above, you can verify your work against the official solutions.

<center>
<a href="../solutions.html" class="btn btn-outline-success" role="button">🔓 Access Lab Solutions</a>
</center>
:::

