# Lab 15: Non-parametric Methods for Hypothesis Testing

## Learning Objectives
* Identify when to use **Non-parametric** tests instead of Parametric tests.
* Conduct a **Wilcoxon Rank-Sum Test** (Mann-Whitney U).
* Conduct a **Wilcoxon Signed-Rank Test** for paired data.
* Understand how "Ranking" data removes the influence of extreme outliers.

---

## Part 1: Parametric vs. Non-parametric
"Parametric" tests (like the t-test) assume your data follows a specific distribution (the Normal Distribution). **Non-parametric** tests make no such assumption. They are often called "Distribution-Free" tests.

### Use a Non-parametric test if:
1. Your sample size is very small (n < 30).
2. Your data is heavily skewed and cannot be fixed.
3. You have major outliers that you cannot remove.
4. Your data is **Ordinal** (e.g., Likert scales: "Satisfied", "Neutral", "Dissatisfied").

---

## Part 2: The Wilcoxon Rank-Sum Test
This is the non-parametric alternative to the **Independent Samples t-test** (Lab 12). Instead of comparing means, it compares the **medians** by ranking all the data points from smallest to largest.



**Scenario:** Are the "Mass" values of Droids and Humans different? (Mass is often heavily skewed by outliers like Jabba the Hutt).
## Wilcoxon Rank-Sum Test: Mass by Species

This test compares the distribution of mass between Humans and Droids. We use this non-parametric test because mass in the Star Wars universe is often skewed by very heavy characters (like Jabba the Hutt).

> **Accessible Logic:** The Wilcoxon test doesn't compare means; it compares the **ranks** of the data. If one species tends to have higher ranks than the other, the test will return a significant result.



```{r}
#| label: tbl-wilcox-mass
#| tbl-cap: "Results of the Wilcoxon Rank-Sum Test comparing mass between Humans and Droids."
#| warning: false
#| message: false

library(tidyverse)
library(broom)

# 1. Filter and clean the data
comparison_data <- starwars |> 
  filter(species %in% c("Human", "Droid")) |>
  drop_na(mass)

# 2. Run the test and save the results
wilcox_results <- wilcox.test(mass ~ species, data = comparison_data)

# 3. Convert results into an accessible table
tidy_wilcox <- tidy(wilcox_results)

knitr::kable(
  tidy_wilcox[, c("statistic", "p.value", "method")],
  col.names = c("W Statistic", "p-value", "Test Method"),
  digits = 4
)
```

---

## Part 3: The Wilcoxon Signed-Rank Test
This is the non-parametric alternative to the **Paired t-test**. Use this when you have "Before" and "After" measurements on the same subjects, but the differences are not normally distributed.

```r
# Example syntax for paired data
# wilcox.test(before_scores, after_scores, paired = TRUE)
```
```{r}
#| label: tbl-paired-wilcox
#| tbl-cap: "Paired Wilcoxon Signed-Rank Test results for character performance."

# Creating mock 'before' and 'after' data for the example
before <- c(80, 75, 90, 85)
after  <- c(82, 78, 88, 90)

tidy_paired <- broom::tidy(wilcox.test(before, after, paired = TRUE))

knitr::kable(
  tidy_paired[, c("statistic", "p.value", "method")],
  col.names = c("V Statistic", "p-value", "Test Method"),
  digits = 4
)
```

---

## Part 4: The Trade-off
While non-parametric tests are safer for "messy" data, they are slightly **less powerful** than parametric tests when the data *is* actually normal. This means they are less likely to find a significant difference if one actually exists. 



---

<div class="dcc-important">
**Lab Task 15:**

1. Pick a numeric variable from your dataset that looks very skewed (not bell-shaped).
2. Create a **Boxplot** of this variable grouped by a category.
3. Run a `wilcox.test()` to compare the two groups.
4. Run a standard `t.test()` on the **same data**.
5. **Compare:** In your Quarto document, compare the p-values. Are they similar? Which test do you trust more for this specific data, and why?
</div>

---

## Pro-Tip: When in Doubt, Rank it Out
If you are presenting data to a stakeholder and you are worried about one or two "crazy" outliers ruining your average, the Wilcoxon test is your best friend. It treats the outlier as just "the highest rank," preventing it from pulling the results in its direction.

---

::: {.callout-caution collapse="true"}
## 📥 Lab Final Submission Instructions (Click to Expand)


### Step 1: Organize Your Folder
Before zipping, verify that your **Files pane** shows the following:

1. You are inside the `Lab_solutions` folder.
2. You have a folder named `Lab01_YourName` (e.g., `Lab01_HenryMendoza`).
3. **Inside** that folder, you have:
    - Your Quarto file (`.qmd`) (e.g.,`Lab01_HenryMendoza.qmd`)
    - Your rendered report (`.html`) (e.g.,`Lab01_HenryMendoza.html`)
    - Your verification screenshot (`.png`) (e.g.,`lab01_task1.png`)

### Step 2: Create the ZIP Archive

1. Open your computer's **File Explorer** (Windows) or **Finder** (Mac).
2. Navigate to your `MAT186_Labs/Lab_solutions/` directory.
3. **Right-click** on your lab folder (e.g., `Lab01_YourName` )
4. Select **Compress** (Mac) or **Send to > Compressed (zipped) folder** (Windows).
5. You will now see a zipped file named (e.g.,`Lab01_YourName.zip`).    
    
### Step 3: Submit to Brightspace

1. Log into **Brightspace** and navigate to **Lab Assignments** (e.g.,`Lab 1: R and RStudio Installation`).
2. Select the correct Lab folder.
3. Upload your **.zip** file and click **Submit**.
:::


---

::: {.callout-tip}
### Done with the Lab?
If you have finished all the tasks above, you can verify your work against the official solutions.

<center>
<a href="../solutions.html" class="btn btn-outline-success" role="button">🔓 Access Lab Solutions</a>
</center>
:::



