# Lab 13: Statistical Inference for Two Proportions

## Learning Objectives
* Compare the "success rates" or percentages of two independent groups.
* State the Null ($H_0$) and Alternative ($H_a$) hypotheses for differences in proportions.
* Use the `prop.test()` function in R for two-sample comparisons.
* Create a **Stacked Bar Chart** to visualize the relative differences between groups.

---

## Introduction
In Lab 11, we looked at one proportion. In the real world, we usually want to compare two. For example: 
* Does Group A have a higher graduation rate than Group B?
* Is a medicine more effective in the treatment group than the control group?

---

## Part 1: Hypotheses for Two Proportions
We are testing the difference between two population proportions ($p_1 - p_2$).

* **Null Hypothesis ($H_0$):** $p_1 = p_2$ (There is no difference between the groups).
* **Alternative Hypothesis ($H_a$):** $p_1 \neq p_2$ (The groups have significantly different proportions).

---

## Part 2: Visualizing with Stacked Bar Charts
When comparing proportions, a standard bar chart can be misleading if the group sizes are different. We use `position = "fill"` to show the **percentage** (0 to 100%) rather than the raw count.

```r
library(tidyverse)

# Comparing 'Success' across two different categories
ggplot(my_data, aes(x = group_variable, fill = success_variable)) +
  geom_bar(position = "fill") +
  labs(title = "Proportional Comparison of Success by Group",
       y = "Proportion (0 to 1)",
       x = "Group Name")
```

```{r}
#| label: fig-gender-tatooine-prop
#| fig-cap: "Stacked bar chart showing the proportion of gender identities for characters from Tatooine versus other homeworlds."
#| fig-alt: "A 100% stacked bar chart with two columns: Tatooine and Other. The y-axis represents proportion from 0 to 1. Both bars are divided into sections representing different genders. The Tatooine bar shows a significantly higher proportion of masculine characters compared to the 'Other' category, which has a slightly more diverse gender distribution."
#| warning: false
#| message: false

library(tidyverse)

# Creating a simplified category: Tatooine vs Other
starwars_props <- starwars |>
  filter(!is.na(gender)) |>
  mutate(location = if_else(homeworld == "Tatooine", "Tatooine", "Other"))

ggplot(starwars_props, aes(x = location, fill = gender)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = "mako") + # High contrast, color-blind friendly palette
  labs(title = "Gender Proportions: Tatooine vs. Other Worlds",
       subtitle = "Standardized to 100% using position = 'fill'",
       x = "Homeworld Category",
       y = "Proportion",
       fill = "Gender")
```


---

## Part 3: Running the Two-Sample Prop Test
In R, we provide the "successes" and the "totals" as vectors within the `prop.test()` function.

**Scenario:** * **Group 1:** 45 people out of 100 succeeded.
* **Group 2:** 30 people out of 100 succeeded.

```{r}
#| label: tbl-two-prop-test
#| tbl-cap: "Two-sample proportion test comparing Group 1 (45/100) and Group 2 (30/100)."

# 1. Perform the test
# x = number of successes (45 and 30)
# n = total sample sizes (100 and 100)
results <- prop.test(x = c(45, 30), n = c(100, 100), correct = FALSE)

# 2. Tidy the results into a data frame
library(broom)
tidy_prop_two <- tidy(results)

# 3. Create an accessible table with descriptive headers
# Note: 'estimate' in a two-sample test contains the proportions for both groups
knitr::kable(
  tidy_prop_two[, c("statistic", "p.value", "conf.low", "conf.high")],
  col.names = c("Chi-squared", "p-value", "Lower 95% CI", "Upper 95% CI"),
  digits = 4
)
```



---

## Part 4: Interpreting the Results

* **p-value:** If the p-value is less than 0.05, we conclude the difference is "Statistically Significant."
* **Sample Estimates:** R will show you the individual proportions ($\hat{p}_1$ and $\hat{p}_2$).
* **Confidence Interval:** This interval estimates the true difference between the two population percentages.

---

<div class="dcc-important">
**Lab Task 13:**

1. Open your Quarto document for Lab 13.
2. Choose two groups from your data and a binary "success" outcome (e.g., "Pass/Fail" or "Yes/No").
3. Create a **Stacked Bar Chart** using `position = "fill"`.
4. Run `prop.test()` to compare the two groups.
5. **Conclusion:** State your p-value and write a one-sentence conclusion. "We [reject/fail to reject] the null hypothesis and conclude there is [a/no] significant difference in the proportions."
</div>

---

## Accessibility Tip: Color and Texture
When creating stacked bar charts, rely on more than just color. Using high-contrast palettes like `scale_fill_brewer(palette = "Set1")` helps students with color blindness distinguish between the "Success" and "Failure" sections of your bars.
