# Lab 12: Statistical Inference for Two Means

## Learning Objectives
* Understand the difference between **Independent** and **Paired** samples.
* State the Null ($H_0$) and Alternative ($H_a$) hypotheses for comparing two groups.
* Conduct an **Independent Samples t-test** in R.
* Visualize the difference between groups using **Boxplots**.

---

## Part 1: Comparing Two Independent Groups
We use an Independent Samples t-test when we want to compare the means of two distinct, unrelated groups.
* **Example:** Do "Humans" have a different average height than "Droids"?
* **Example:** Is the average house price in Poughkeepsie different from the average price in Fishkill?



---

## Part 2: Hypotheses for Two Means
In this test, we are looking at the **difference** between the two population means ($\mu_1 - \mu_2$).

* **Null Hypothesis ($H_0$):** $\mu_1 = \mu_2$ (The means are the same; the difference is zero).
* **Alternative Hypothesis ($H_a$):** $\mu_1 \neq \mu_2$ (The means are significantly different).

---

## Part 3: Visualizing the Comparison
Before running the math, we should always visualize the two groups side-by-side using a **Boxplot**.

```{r}
#| label: fig-height-human-droid
#| fig-cap: "Comparison of height distributions between Humans and Droids."
#| fig-alt: "A side-by-side boxplot comparing heights of Droids and Humans. The Droid distribution shows much greater variability, with a box and whiskers that span a wider range of heights. The Human distribution is more compact, with a median height slightly higher than the Droids. Several outliers are visible for the Droids, representing exceptionally short and tall individuals."
#| warning: false
#| message: false

library(tidyverse)

# Comparing heights of Humans and Droids
comparison_data <- starwars |> 
  filter(species %in% c("Human", "Droid")) |>
  drop_na(height)

ggplot(comparison_data, aes(x = species, y = height, fill = species)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set2") + # Better color contrast for accessibility
  labs(title = "Height Comparison: Humans vs. Droids",
       x = "Species",
       y = "Height (cm)")
```



---

## Part 4: Running the t-test in R
In R, we use the "formula" syntax: `y ~ x` (where `y` is the numeric value and `x` is the grouping category).

```{r}
#| label: tbl-two-sample-t-test
#| tbl-cap: "Independent samples t-test comparing the mean heights of Humans and Droids."

# 1. Conduct the Independent Samples t-test
t_test_results <- t.test(height ~ species, data = comparison_data)

# 2. Tidy the results into a data frame
library(broom)
tidy_t_test <- tidy(t_test_results)

# 3. Create an accessible table with descriptive headers
knitr::kable(
  tidy_t_test[, c("estimate", "statistic", "p.value", "conf.low", "conf.high")],
  col.names = c("Difference in Means", "t-statistic", "p-value", "Lower 95% CI", "Upper 95% CI"),
  digits = 3
)
```

### How to interpret the output:
* **p-value:** If p < 0.05, we conclude that the average heights of Humans and Droids are significantly different.
* **95% Confidence Interval:** This gives the range of the *difference* between the two means. If the interval contains **0**, the results are usually not significant.

---

<div class="dcc-important">
**Lab Task 12:**

1. Pick a dataset that has at least one **categorical variable** with two groups (e.g., Male/Female, Treatment/Control, Yes/No).
2. Create a **Boxplot** to visually compare a numeric variable across those two groups.
3. State your Null and Alternative hypotheses.
4. Run the `t.test()` using the formula syntax.
5. **The Conclusion:** Report your p-value and state whether you reject or fail to reject the null hypothesis. Does the data suggest the groups are truly different?
</div>

---

## Accessibility Tip: High-Contrast Visuals
When using the `fill = species` aesthetic in your plots, remember that some students may have color vision deficiencies. Use `scale_fill_viridis_d()` to ensure your group comparisons are accessible to everyone.

---

::: {.callout-caution collapse="true"}
## 📥 Lab Final Submission Instructions (Click to Expand)


### Step 1: Organize Your Folder
Before zipping, verify that your **Files pane** shows the following:

1. You are inside the `Lab_solutions` folder.
2. You have a folder named `Lab01_YourName` (e.g., `Lab01_HenryMendoza`).
3. **Inside** that folder, you have:
    - Your Quarto file (`.qmd`) (e.g.,`Lab01_HenryMendoza.qmd`)
    - Your rendered report (`.html`) (e.g.,`Lab01_HenryMendoza.html`)
    - Your verification screenshot (`.png`) (e.g.,`lab01_task1.png`)

### Step 2: Create the ZIP Archive

1. Open your computer's **File Explorer** (Windows) or **Finder** (Mac).
2. Navigate to your `MAT186_Labs/Lab_solutions/` directory.
3. **Right-click** on your lab folder (e.g., `Lab01_YourName` )
4. Select **Compress** (Mac) or **Send to > Compressed (zipped) folder** (Windows).
5. You will now see a zipped file named (e.g.,`Lab01_YourName.zip`).    
    
### Step 3: Submit to Brightspace

1. Log into **Brightspace** and navigate to **Lab Assignments** (e.g.,`Lab 1: R and RStudio Installation`).
2. Select the correct Lab folder.
3. Upload your **.zip** file and click **Submit**.
:::


---

::: {.callout-tip}
### Done with the Lab?
If you have finished all the tasks above, you can verify your work against the official solutions.

<center>
<a href="../solutions.html" class="btn btn-outline-success" role="button">🔓 Access Lab Solutions</a>
</center>
:::

